---
title: "Getting started with msomgom"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with msomgom}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
have_jags <- requireNamespace("rjags", quietly = TRUE) && requireNamespace("dclone", quietly = TRUE)
have_fancyfx <- requireNamespace("fancyfx", quietly = TRUE)
```

msomgom fits a dynamic (multi-season, colonization/persistence) occupancy
model - following MacKenzie et al. (2003), an extension of the single-season
detection model of MacKenzie et al. (2002) - to cetacean vessel-survey
sightings, via JAGS (Plummer 2003). It was originally built around Bay of
Fundy North Atlantic right whale survey data from the NARWC Sightings
Database, then generalized so the study area, date range, species, and MCMC
settings are all driven by a config rather than hardcoded. See the package
README for full citations and setup instructions (JAGS itself has to be
installed separately, and needs to be version 4.x).

This vignette walks through the whole pipeline using synthetic data, since
the real NARWC survey CSV has its own data-sharing terms and isn't included
in the package. The mock data generator, `generate_mock_data()`, is grounded
in the NARWC handbook's actual field codes rather than arbitrary values, so
it's a real exercise of the pipeline's filtering logic, not just a
happy-path stub.

```{r setup}
library(msomgom)
```

## 1. Configure a run

Configs are YAML files, one per run, so different study areas/species/MCMC
settings can be kept side by side. `generate_config()` writes one
programmatically rather than requiring you to hand-edit YAML; its defaults
reproduce the pipeline's original Bay of Fundy / right whale (RIWH) settings.
Here we override the MCMC settings to be small and fast, since this vignette
needs to actually run:

```{r configure}
work_dir <- tempfile("msomgom-vignette-")
dir.create(work_dir)

config_path <- generate_config(
  "vignette_demo",
  configs_dir = file.path(work_dir, "configs"),
  project_dir = work_dir,
  data_file = "data/mock_survey_data.csv",
  output_dir = "output",
  beg_year = 2018, end_year = 2019,
  beg_month = 8, end_month = 9,
  n_chains = 2, n_adapt = 100, n_burn = 100, n_iter = 200, thin = 1
)
config_path
```

See the comments in a generated config, or `?generate_config`, for what every
field means: survey vessel/`FILEID` filters, date range, season boundaries,
the study-area polygon, species, JAGS/MCMC settings, evaluation, and cleanup.

## 2. Generate mock survey data

```{r mock-data}
generate_mock_data(config_path)
```

## 3. Run the pipeline

`run_occupancy_model()` runs every stage in order: `prep_survey_data()` (load
+ clean the survey CSV) -> `build_detection_arrays()` (spatial grid +
detection-history arrays) -> `fit_occupancy_model()` (fit the model) ->
`evaluate_occupancy_model()` (convergence diagnostics) -> `cleanup_outputs()`
(only if configured to).

```{r run, eval = have_jags}
result <- run_occupancy_model(config_path)
result$fit
```

```{r run-noeval, eval = !have_jags, echo = FALSE, results = "asis"}
cat("_(This chunk needs a working JAGS/rjags/dclone install to actually run;",
    "skipped when building this vignette without one. See README.md for JAGS setup.)_")
```

`result$fit` is the fitted `mcmc.list`. `result$evaluation` (see next
section) has already been computed and printed above, since
`run_occupancy_model()` calls `evaluate_occupancy_model()` by default.

### Debugging a run that isn't working

Before reaching for R's general-purpose debugger, `diagnose_pipeline()` runs
the data-prep and grid-building stages on their own (never JAGS) and reports
the most common reasons a run fails, hangs, or produces meaningless output -
as a plain PASS/WARN/FAIL list:

```r
diagnose_pipeline(config_path)
```

If the problem is a genuine crash rather than something `diagnose_pipeline()`
flags, R's own debugging tools work on any msomgom function the same way
they'd work on your own code:

```r
# see the exact call stack right after an error
traceback()

# step through a function line by line ("n" for next line, "c" to finish)
debugonce(prep_survey_data)
prep_survey_data(load_config(config_path), verbose = TRUE)

# drop into the debugger at *any* uncaught error, for the rest of the session
options(error = recover)
run_occupancy_model(config_path)
options(error = NULL) # turn it back off when done
```

`prep_survey_data(..., verbose = TRUE)` is worth calling on its own too - it
prints the record count after every filter step (platform, `FILEID` prefix,
date range), which is usually the fastest way to see exactly where a filter
is dropping more than expected.

## 4. Evaluate the fit

`evaluate_occupancy_model()` reports a posterior parameter summary table,
Gelman-Rubin Rhat and effective sample size per parameter (flagging likely
non-convergence or a noisy posterior), and saves trace/density plots. It also
works standalone against a previously saved result, without re-fitting:

```{r evaluate, eval = have_jags}
result$evaluation$parameters
```

With the tiny MCMC settings used in this vignette (100 adapt / 100 burn-in /
200 iterations, for speed), don't expect real convergence - a real run should
use far more iterations, e.g. the defaults in `generate_config()`.

## 5. Environmental covariates (optional)

`whale.mod` can put environmental covariates on detection probability
(already the case for day-of-year, sea state, and effort), and optionally on
initial occupancy (`psi`), persistence (`phi`), and colonization (`gamma`)
too - each process independently, with zero or more named covariates chosen
per config. A process with none configured stays intercept-only.

`average_covariates()` preps covariate data onto this pipeline's hex grid, at
either the occupancy model's own season/year windows or arbitrary custom
ones. It accepts an `sf` point object tagged with YEAR/MONTH/DAY from any of
a few sources that all share that same shape:
[`datamatch::accessCopernicus()`](https://github.com/chross22/datamatch), which
pulls live from the Copernicus Marine Service; local daily NetCDF files via
this package's own `load_covariate_netcdf()`; or either of those further
enriched by [`derivoce`](https://github.com/chross22/derivoce)'s derived
covariates (spatial/temporal gradients, distance to shore/front/isobath,
lags, integrals, ...), which consume and return the same shape:

```r
# not run in this vignette, since it needs network access and Copernicus
# credentials - see datamatch's README for setup. Either package is optional
# (Suggests), installed the same way as this one:
#   devtools::install_github("chross22/datamatch")
#   devtools::install_github("chross22/derivoce")
env_dat <- datamatch::accessCopernicus(
  vars = "SST", years = 2018:2020, months = 1:12,
  bounding_box = list(xmin = -70, xmax = -66, ymin = 41, ymax = 44)
)
env_dat <- derivoce::horizontal_gradient(env_dat, "SST") # adds an SST_grad column

windows <- season_windows_from_config(config)
sst_avg <- average_covariates(env_dat, arrays$area_grid_sf, windows)
```

The rest of this section uses a synthetic covariate instead, so the vignette
renders offline without those credentials - the mechanics past this point
(config wiring, `occ_covariates`) are identical either way:

```{r covariates, eval = have_jags}
config <- load_config(config_path)
prep <- prep_survey_data(config)
arrays <- build_detection_arrays(prep$tmpdat, prep$season_info, config)

# a synthetic covariate here, standing in for a real one from
# load_covariate_netcdf(), datamatch::accessCopernicus(), or derivoce (see above)
set.seed(1)
windows <- season_windows_from_config(config)
fake_sst <- matrix(
  rnorm(arrays$num_cells * nrow(windows), mean = 15, sd = 2),
  nrow = arrays$num_cells, dimnames = list(NULL, windows$label)
)

config_cov <- generate_config(
  "vignette_demo_covariates",
  configs_dir = file.path(work_dir, "configs"),
  project_dir = work_dir,
  data_file = "data/mock_survey_data.csv",
  output_dir = "output_cov",
  beg_year = 2018, end_year = 2019,
  beg_month = 8, end_month = 9,
  n_chains = 2, n_adapt = 100, n_burn = 100, n_iter = 200, thin = 1,
  covariates_psi = c("sst")
)

result_cov <- run_occupancy_model(config_cov, occ_covariates = list(sst = fake_sst))
result_cov$evaluation$parameters$parameter # now includes mu.b.cov (the sst effect on psi)
```

Since JAGS can't compile a covariate loop of length zero, the model's BUGS
code is generated per run (`build_whale_model_code()`) rather than being one
static model - a config with no covariates produces exactly the original
intercept-only model, and each process's covariate block is included only
when that process actually has covariates configured.

### Plotting the covariate's effect

`result_cov$fit` is tagged `msomgom_fit`, which carries what
[`fancyfx`](https://github.com/chross22/fancyfx) needs to plot `sst`'s fitted
effect - the same way it plots an `mgcv::gam`'s partial effects or any other
model's predictions, via its `effect_estimates()` generic. `fancyfx` is
optional (not installed with msomgom); this section is skipped if it isn't
available.

```{r effects, eval = have_jags && have_fancyfx}
library(fancyfx)

effect_estimates(result_cov$fit, "sst")
```

```{r effects-noeval, eval = !(have_jags && have_fancyfx), echo = FALSE, results = "asis"}
cat("*(Skipped: needs both a JAGS install and the fancyfx package - not run",
    "when building this vignette without them.)*")
```

The full plot stacks a rug of the raw covariate data above the effect curve:

```{r effects-plot, eval = have_jags && have_fancyfx, fig.width = 6, fig.height = 4}
rug_dat <- data.frame(sst = as.vector(fake_sst))
plotEffects(result_cov$fit, rug_dat, "sst", xlab = "SST")
```

## Cleanup

```{r cleanup, include = FALSE}
unlink(work_dir, recursive = TRUE)
```
